---
title: "Functional Analysis of Heart Rate Dynamics"
author: "Jenna Lu"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

# Overview

Physiological responses are dynamic processes that evolve over time rather than static measurements.

This notebook explores heart rate trajectories as functional data and investigates how cardiac responses vary across experimental conditions in the WESAD dataset.

The goals are:

1. Visualize heart rate trajectories
2. Map cardiac responses to experimental conditions
3. Smooth physiological curves
4. Compare temporal patterns across conditions


# Libraries

```{r setup, message=FALSE, warning=FALSE}

library(tidyverse)
library(here)
library(fda)
library(data.table)
```


# Load Heart Rate Data

```{r}

hr_data <- read_csv(
  here(
    "data",
    "processed",
    "S2_heart_rate.csv"
  )
)

glimpse(hr_data)

```


# Load Condition Labels

The original ECG data contains experimental condition labels.

```{r}

ecg_data <- read_csv(
  here(
    "data",
    "processed",
    "S2_ecg_filtered.csv"
  )
)


condition_data <- ecg_data %>%
  dplyr::select(
    time,
    label
  )


```


# Match Heart Rate to Conditions

Heart rate is measured at detected beat locations, while labels are sampled at the ECG frequency.

The nearest time point is used to associate physiological states with heart rate measurements.

```{r}

hr_condition <- data.table(hr_data)

condition_table <- data.table(condition_data)

hr_condition <- condition_table[
  hr_condition,
  on = .(
    time = time
  ),
  roll = "nearest"
]

hr_condition <- as_tibble(hr_condition)


head(hr_condition)

```


# Condition Summary

```{r}

hr_condition %>%
  dplyr::group_by(label) %>%
  dplyr::summarise(
    mean_hr = mean(heart_rate, na.rm=TRUE),
    sd_hr = sd(heart_rate, na.rm=TRUE),
    n=dplyr::n()
  )

```


# Visualize Heart Rate Trajectory

```{r}

ggplot(
  hr_condition,
  aes(
    x=time,
    y=heart_rate
  )
) +
  geom_line(alpha=0.5) +
  labs(
    title="Heart Rate Trajectory Across Recording",
    x="Time (seconds)",
    y="Heart Rate (bpm)"
  ) +
  theme_minimal()

```


# Smooth Heart Rate Function

Physiological signals contain short-term variability. Functional data analysis represents the underlying trajectory using smooth basis functions.

```{r}

time_grid <- hr_condition$time

basis <- create.bspline.basis(
  rangeval = range(time_grid),
  nbasis = 25
)


hr_fd <- Data2fd(
  argvals = time_grid,
  y = hr_condition$heart_rate,
  basisobj = basis
)


hr_fd

```


# Plot Functional Representation

```{r}

plot(
  hr_fd,
  main="Smoothed Heart Rate Functional Representation",
  xlab="Time (seconds)",
  ylab="Heart Rate"
)

```


# Functional Derivative

The first derivative describes the rate of change of heart rate over time.

```{r}
hr_velocity <- fda::deriv.fd(
  hr_fd,
  Lfdobj = 1
)

plot(
  hr_velocity,
  main="Rate of Change of Heart Rate",
  xlab="Time",
  ylab="dHR/dt"
)

```


# Summary

This notebook represents physiological signals as continuous functions rather than isolated measurements.

Functional representations allow future modeling of:

- stress-response trajectories
- individual variability
- longitudinal physiological patterns
- condition-specific cardiac dynamics