---
title: "Statistical Modeling of Physiological Responses"
author: "Jenna Lu"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

# Overview

Physiological measurements often exhibit systematic changes across experimental conditions.

This notebook applies statistical modeling approaches to quantify associations between experimental state and heart rate response.

The goals are:

1. Compare heart rate across physiological conditions
2. Model condition-associated changes
3. Quantify uncertainty in estimated effects
4. Evaluate model assumptions


# Libraries

```{r setup, message=FALSE, warning=FALSE}

library(tidyverse)
library(here)
library(broom)

```


# Load Heart Rate Data

```{r}

hr_data <- read_csv(
  here(
    "data",
    "processed",
    "S2_heart_rate.csv"
  )
)

head(hr_data)

```


# Load Condition Information

```{r}

ecg_data <- read_csv(
  here(
    "data",
    "processed",
    "S2_ecg_filtered.csv"
  )
)

condition_data <- ecg_data %>%
  dplyr::select(
    time,
    label
  )


```


# Align Conditions

```{r}

hr_condition <- hr_data %>%
  mutate(
    label = condition_data$label[
      findInterval(
        time,
        condition_data$time
      )
    ]
  )


hr_condition <- hr_condition %>%
  filter(!is.na(label))


```


# Convert Conditions to Factors

```{r}

hr_condition <- hr_condition %>%
  mutate(
    condition = factor(label)
  )


table(hr_condition$condition)

```


# Exploratory Comparison

```{r}

hr_condition %>%
  ggplot(
    aes(
      x=condition,
      y=heart_rate
    )
  ) +
  geom_boxplot() +
  labs(
    title="Heart Rate Across Experimental Conditions",
    x="Condition",
    y="Heart Rate (bpm)"
  ) +
  theme_minimal()

```


# Linear Model

We model heart rate as a function of experimental condition.

```{r}

hr_model <- lm(
  heart_rate ~ condition,
  data = hr_condition
)


summary(hr_model)

```


# Estimated Effects

```{r}

tidy(hr_model)

```


# Model Diagnostics

```{r}

par(mfrow=c(2,2))

plot(hr_model)

```


# Interpretation

The regression model estimates average changes in heart rate associated with each physiological condition relative to the reference state.

Future extensions include:

- mixed-effects models across participants
- Bayesian hierarchical models
- functional regression approaches
- machine learning prediction of physiological states


# Summary

This analysis demonstrates statistical modeling of dynamic physiological signals and provides a foundation for more advanced longitudinal and hierarchical modeling.