---
title: "Machine Learning Prediction of Physiological States"
author: "Jenna Lu"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

# Overview

Physiological signals contain patterns that may allow prediction of human state.

This notebook develops predictive models using heart rate features extracted from ECG recordings.

The goals are:

1. Engineer physiological predictors
2. Define stress classification targets
3. Train statistical learning models
4. Evaluate predictive performance


# Libraries

```{r setup, message=FALSE, warning=FALSE}

library(tidyverse)
library(here)
library(tidymodels)
library(pROC)

```


# Load Heart Rate Data

```{r}

hr_data <- read_csv(
  here(
    "data",
    "processed",
    "S2_heart_rate.csv"
  )
)

```


# Load Conditions

```{r}

ecg_data <- read_csv(
  here(
    "data",
    "processed",
    "S2_ecg_filtered.csv"
  )
)


condition_data <- ecg_data %>%
  dplyr::select(
    time,
    label
  )


```


# Align Conditions

```{r}

hr_condition <- hr_data %>%
  mutate(
    label = condition_data$label[
      findInterval(
        time,
        condition_data$time
      )
    ]
  ) %>%
  filter(!is.na(label))


```


# Create Prediction Target

Stress conditions in WESAD correspond to label 2.

We create a binary classification problem.

```{r}

model_data <- hr_condition %>%
  mutate(
    stress = ifelse(
      label == 2,
      "Stress",
      "Non-Stress"
    ),
    stress = factor(stress)
  )


table(model_data$stress)

```


# Feature Engineering

We summarize heart rate into short temporal windows.

```{r}

features <- model_data %>%
  mutate(
    window = floor(time / 60)
  ) %>%
  group_by(window, stress) %>%
  summarise(
    mean_hr = mean(heart_rate),
    sd_hr = sd(heart_rate),
    max_hr = max(heart_rate),
    min_hr = min(heart_rate),
    .groups="drop"
  )


head(features)

```


# Train/Test Split

```{r}

set.seed(123)

split <- initial_split(
  features,
  prop = 0.75,
  strata = stress
)

train <- training(split)

test <- testing(split)


```


# Logistic Regression Model

```{r}

log_model <- logistic_reg() %>%
  set_engine("glm") %>%
  fit(
    stress ~ mean_hr + sd_hr + max_hr + min_hr,
    data=train
  )


log_model

```


# Predictions

```{r}

predictions <- predict(
  log_model,
  test,
  type="prob"
) %>%
  bind_cols(test)


head(predictions)

```


# ROC Curve

```{r}

roc_curve <- roc(
  predictions$stress,
  predictions$.pred_Stress
)


auc(roc_curve)

```

# Cross-Validation

```{r}

```{r}

set.seed(123)

folds <- vfold_cv(
  train,
  v = 5,
  strata = stress
)


log_recipe <- recipe(
  stress ~ mean_hr + sd_hr + max_hr + min_hr,
  data = train
)


log_workflow <- workflow() %>%
  add_recipe(log_recipe) %>%
  add_model(
    logistic_reg() %>%
      set_engine("glm")
  )


cv_results <- fit_resamples(
  log_workflow,
  resamples = folds,
  metrics = metric_set(
    roc_auc,
    accuracy
  )
)


collect_metrics(cv_results)


```


# ROC Visualization

```{r}

plot(
  roc_curve,
  main="ROC Curve for Stress Classification"
)

```


# Summary

This notebook demonstrates predictive modeling of physiological state using cardiac features.

Future extensions include:

- random forest models
- Bayesian classification
- mixed-effects machine learning
- multi-subject validation
- multimodal EEG and ECG integration