---
title: "Bayesian Hierarchical Modeling of Physiological Responses"
author: "Jenna Lu"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

# Overview

Physiological responses vary across individuals. 
A hierarchical modeling framework allows population-level effects to be estimated while accounting for participant-specific variability.

This notebook develops a Bayesian mixed-effects model for heart rate responses across participants in the WESAD dataset.

The goals are:

1. Construct a multi-subject physiological dataset
2. Model stress-related heart rate changes
3. Estimate uncertainty using Bayesian inference
4. Quantify individual differences in baseline physiology


# Libraries

```{r setup, message=FALSE, warning=FALSE}

library(tidyverse)
library(here)
library(brms)
library(bayesplot)
library(posterior)

```


# Load Multi-Subject Data

The feature dataset generated in the previous analysis contains heart rate summaries across participants.

```{r}

features <- read_csv(
  here(
    "data",
    "processed",
    "multisubject_features.csv"
  )
)

head(features)

```


# Create Stress Outcome

Stress corresponds to WESAD label 2.

```{r}

model_data <- features %>%
  mutate(
    stress = ifelse(
      label == 2,
      1,
      0
    )
  )


model_data %>%
  count(stress)

```


# Explore Participant Variability

Participants exhibit different baseline heart rate distributions.

```{r}

model_data %>%
  ggplot(
    aes(
      x=subject,
      y=mean_hr
    )
  ) +
  geom_boxplot() +
  labs(
    title="Heart Rate Variation Across Participants",
    x="Participant",
    y="Mean Heart Rate"
  ) +
  theme_minimal()

```


# Bayesian Hierarchical Model

We model mean heart rate as a function of stress while allowing each participant to have their own baseline.

```{r message=FALSE}

bayes_model <- brm(

  mean_hr ~ stress + 
    (1 | subject),

  data=model_data,

  family=gaussian(),

  prior=c(

    prior(
      normal(0,10),
      class="b"
    ),

    prior(
      exponential(1),
      class="sd"
    )

  ),

  chains=4,
  iter=2000,
  warmup=1000,
  seed=123

)


```


# Model Summary

```{r}

summary(bayes_model)

```


# Posterior Distributions

Bayesian inference provides full posterior distributions rather than only point estimates.

```{r}

plot(
  bayes_model
)

```


# Stress Effect Estimate

The stress coefficient represents the estimated change in heart rate associated with stress.

```{r}

posterior_samples <- posterior_samples(
  bayes_model
)


posterior_samples %>%
  select(
    b_stress
  ) %>%
  summary()

```


# Credible Interval Visualization

```{r}

mcmc_areas(
  posterior_samples,
  pars="b_stress"
)

```


# Participant Random Effects

Individual baseline differences are captured through participant-level effects.

```{r}

ranef(bayes_model)

```


# Posterior Predictive Checking

We evaluate whether simulated data from the model resemble observed physiological patterns.

```{r}

pp_check(
  bayes_model
)

```


# Summary

This analysis demonstrates Bayesian hierarchical modeling of physiological signals.

The model:

- estimates population-level stress effects
- accounts for individual physiological differences
- quantifies uncertainty through posterior distributions
- provides a framework for personalized biomedical modeling


Future extensions include:

- Bayesian functional regression
- time-varying hierarchical models
- multimodal EEG and ECG integration
- personalized physiological prediction