---
title: "Multi-Subject Validation of Physiological Responses"
author: "Jenna Lu"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

# Overview

Physiological responses vary substantially between individuals. 

This notebook extends the previous analyses from a single participant to multiple participants in the WESAD dataset.

The goals are:

1. Process heart rate features across multiple subjects
2. Evaluate subject-level variability
3. Compare physiological responses across conditions
4. Fit hierarchical statistical models accounting for repeated measurements


# Libraries

```{r setup, message=FALSE, warning=FALSE}

library(tidyverse)
library(here)
library(lme4)
library(broom.mixed)
```


# Load Subject-Level ECG Data

The converted WESAD files contain ECG recordings for each participant.

We use the processed ECG files generated during preprocessing.

```{r}

files <- list.files(
  here("data", "processed"),
  pattern = "_ecg.csv$",
  full.names = TRUE
)


files

```


# Extract Subject Information

```{r}

subject_data <- tibble(
  file = files
) %>%
  mutate(
    subject = str_extract(
      basename(file),
      "S[0-9]+"
    )
  )


subject_data

```


# Load Heart Rate Data

The heart rate files generated previously are combined across participants.

```{r}

hr_files <- list.files(
  here("data", "processed"),
  pattern = "_heart_rate.csv$",
  full.names = TRUE
)


hr_all <- map_dfr(
  hr_files,
  function(file){

    subject <- str_extract(
      basename(file),
      "S[0-9]+"
    )

    read_csv(file) %>%
      mutate(
        subject = subject
      )

  }
)


glimpse(hr_all)

```


# Summary of Participants

```{r}

hr_all %>%
  count(subject)

```


# Load Condition Labels

Because heart rate occurs at detected beat locations, conditions are matched using the original ECG timestamps.

```{r}

ecg_files <- list.files(
  here("data", "processed"),
  pattern="_ecg_filtered.csv$",
  full.names=TRUE
)


```


# Create Subject-Level Condition Data

```{r}

condition_all <- map_dfr(
  ecg_files,
  function(file){

    subject <- str_extract(
      basename(file),
      "S[0-9]+"
    )


    read_csv(file) %>%
      select(time, label) %>%
      mutate(
        subject = subject
      )

  }
)


head(condition_all)

```


# Align Heart Rate With Conditions

```{r}

hr_condition_all <- hr_all %>%
  group_by(subject) %>%
  group_modify(
    ~{

      subject_id <- .y$subject

      labels <- condition_all %>%
        filter(subject == subject_id) %>%
        arrange(time)

      .x %>%
        mutate(
          label = labels$label[
            findInterval(
              time,
              labels$time
            )
          ]
        )

    }
  ) %>%
  ungroup()
```


# Condition Summary Across Subjects

```{r}

hr_condition_all %>%
  group_by(label) %>%
  summarise(
    mean_hr = mean(heart_rate),
    sd_hr = sd(heart_rate),
    n = n()
  )

```


# Subject Variability

Individuals may have different baseline heart rates.

```{r}

hr_condition_all %>%
  ggplot(
    aes(
      x=subject,
      y=heart_rate
    )
  ) +
  geom_boxplot() +
  labs(
    title="Heart Rate Distribution Across Participants",
    x="Subject",
    y="Heart Rate (bpm)"
  ) +
  theme_minimal()

```


# Mixed Effects Model

Repeated measurements from the same participant are not independent.

A mixed-effects model accounts for subject-specific baseline differences.

```{r}

hr_model <- lmer(
  heart_rate ~ factor(label) + (1|subject),
  data = hr_condition_all
)


summary(hr_model)

```


# Fixed Effects Estimates

```{r}

tidy(
  hr_model,
  effects="fixed"
)

```


# Random Effects

```{r}

ranef(hr_model)

```


# Interpretation

The mixed-effects model evaluates whether physiological conditions are associated with changes in heart rate while accounting for variability between individuals.

This framework provides a foundation for:

- hierarchical Bayesian models
- personalized physiological modeling
- longitudinal biomedical analysis
- multimodal physiological prediction


# Summary

This analysis extends physiological modeling from an individual participant to a population-level framework.

By incorporating subject-specific variability, the pipeline better reflects real-world biomedical data where repeated measurements are collected from heterogeneous individuals.